---
title: "A Geographic Exploration of Mental Healthcare"
author: "Lucas Bernstein, Vera Chappell"
date: "Add a date"
output:
  rmdformats::material:
    highlight: "kate"
---

```{r setup, include = FALSE}
# Set code chunk defaults 
# Consider adding `message = FALSE` option
knitr::opts_chunk$set(echo = FALSE) 
                      
# Set R environment options
options(knitr.kable.NA = '')

# Load packages
library(tidyverse)

```

# Intro

On February 19, 2023, the U.S. Department of Education announced that they were awarding $188 million to increase access to mental health and wellness services at schools across the country. It’s a decision that comes at a time when mental health issues are more in the spotlight than ever – more than one in three high school students experienced poor mental health during the pandemic. Clearly, this is a national issue revealing itself at an alarming rate, which is why it is more crucial than ever to recognize what might be behind such a systemic problem and identify ways to resolve it.

One possible factor we chose to explore is geographic location. Given that funding for mental health related issues [differs widely by state](https://rehabs.com/explore/mental-health-spending-by-state-across-the-us/), we were curious to explore how these differences in resource allocation might affect how mental healthcare patients are diagnosed and what type of treatment they are given.

Another is the co-occurrence, or comorbidity, of multiple mental health or substance abuse disorders in one person. Not only are over 50% of people diagnosed with one psychiatric disorder [also diagnosed with another at some point in their lives](https://www.colorado.edu/today/2022/05/10/multiple-diagnoses-are-norm-mental-illness-new-genetic-study-explains-why), but those with co-occurring disorders are [often recommended different types of treatment](https://americanaddictioncenters.org/co-occurring-disorders), fundamentally changing their outcomes. For these reasons and more, understanding both the causes and effects of comorbid mental health diagnoses are crucial to fully interpreting the psychiatric situation in the United States.

On the topic of treatment, another way in which mental healthcare patients’ experiences differ is in the services they have access to and receive. [Nearly a third of American adults don’t receive necessary mental health treatment](https://mhanational.org/issues/state-mental-health-america), which already raises possible connections to state-by-state funding and receiving multiple diagnoses. Additionally, while community-based mental health services are [on the rise worldwide](https://www.who.int/news-room/feature-stories/detail/community-based-mental-health-services-using-a-rights-based-approach), are they as effective as traditional inpatient or justice service-based treatments?

By examining geography, comorbidity, and availability of services in an interconnected way, we hoped to answer these questions and provide some context to a clearly complex situation.


## Our Topic

For this project, we chose to continue to explore mental health as we did in our Shiny app. However, while our Shiny app focused exclusively on students, we wanted to expand our target population to include all adults ages 18-39. Additionally, our previous project explored how demographic factors such as age, race, and education affected mental health diagnoses and outcomes; while we are still looking at diagnoses and outcomes here, we wanted to see how incorporating spatial data - specifically state-by-state breakdowns - might lead us to additional insights in these areas.  


## Our Data

Our data comes from the Substance Abuse and Mental Health Services Administration's (SAMHSA) Mental Health Client Level Data for 2020. [According to SAMHSA](https://www.samhsa.gov/data/data-we-collect/mh-cld-mental-health-client-level-data), the data "are for individuals receiving mental health treatment services provided or funded through state mental health agencies". Because of this overall population represented in the dataset, we narrowed it down further to only include those who received at least one primary diagnosis from mental healthcare providers (since it is already an unrepresentative sample in that sense), as well as our age range of 18-39.


# Exploring the Data

## Mental Health Comorbidities by State

To begin, we created [an interactive choropleth](https://lucasbernstein.shinyapps.io/pug_blog_comorbidity/) exploring how often clients who are diagnosed with one condition are also given additional diagnoses, broken down by US state. (To recap: the term comorbidity, according to the NIH, “describes two or more disorders or illnesses occurring in the same person.”) For each diagnosis you can select, the comorbidity rate for each state is the percentage of people primarily diagnosed with the selected condition who were also given at least one additional diagnosis.

```{r comorbidities-app, out.width = "100%", out.extra = 'data-external="1" frameborder="no"', fig.cap = "[Mental Health Comorbidities Shiny App](https://lucasbernstein.shinyapps.io/pug_blog_comorbidity/)"}
knitr::include_app("https://lucasbernstein.shinyapps.io/pug_blog_comorbidity/",
                   height = "500px")
```

Interestingly, no real geographic patterns emerged. For example, we thought that states in the Deep South might have lower comorbidity rates because of lower mental health funding and general stigma surrounding mental healthcare, but with the exception of Alabama and Florida these states tended to be around the national average for each condition. However, we did have a few state-by-state observations:

* Certain states have higher than average comorbidity rates regardless of primary diagnosis (e.g. Minnesota, Washington, Virginia), often approaching 80%; however, these states often have low to average comorbidity rates when the primary diagnosis is delirium/dementia, substance use disorders, or developmental disorders.
* Some disorders, like depressive, bipolar, and conduct, appear to have higher comorbidity rates nationwide. This could be due to the nature of the conditions themselves, or how people who are initially diagnosed with these conditions are further evaluated.
* California has a low comorbidity rate across most primary diagnoses even compared to others with similar populations. Arkansas' low comorbidity rate is at an even larger extreme (one of the lowest for every primary diagnosis).

Additionally, we created [a table](https://vchappell.shinyapps.io/ServicesTable/) observing the different health services used by participants which can be filtered by Gender, Employment Status, and Education level, broken down at a state level.

To help understand this table, the different abbreviated services are as follows :

- SPHSERVICE -> State-psychiatric hospital
- CMPSERVICE -> SMHA-funded/operated community-based program
- OPISERVICE -> Other psychiatric inpatient center
- RTCSERVICE -> Residential treatment center
- IJSSERVICE -> Institution under the justice system

```{r services-app, out.width = "100%", out.extra = 'data-external="1" frameborder="no"', fig.cap = "[Mental Health Services Shiny App](https://vchappell.shinyapps.io/ServicesTable/)"}
knitr::include_app("https://vchappell.shinyapps.io/ServicesTable/",
                   height = "500px")
```
The display of this data will eventually be reformatted to display a proportion of the whole such that results are more easily understood, but some preliminary results are:

* For nearly every state, there is an inverse relationship between those served by residential treatment centers and institutions under the justice system (i.e., if more are served by an IJS, less are served by a RTC and vice-versa.)

* Community-based program services are the most utilized services in each state compared to the other services documented.

* In nearly all states, there is a higher sum of men seen in state-psychiatric hospitals, other psychiatric inpatient centers, and institutions under the justice system than women. 

# Closing Remarks 

Some things to consider with the data we have used are the method of collection and time of collection. Via the [codebook for the SAMHSA data](https://www.datafiles.samhsa.gov/sites/default/files/field-uploads-protected/studies/MH-CLD-2020/MH-CLD-2020-datasets/MH-CLD-2020-DS0001/MH-CLD-2020-DS0001-info/MH-CLD-2020-DS0001-info-codebook.pdf), “The scope of facilities from which the data is collected is affected by differences in state licensure, certification, accreditation, and disbursement of public funds.” Along with what has been noted via the codebook, there is no data from Maine, New Hampshire, West Virginia, and Maryland, and data from these states may or may not significantly change our findings or connections. 

Furthermore, this data was collected in 2020, which is when the Covid-19 pandemic emerged. This may have altered the data in some way or not, but understanding the context within which data has been obtained is important to interpret the results.

To conclude, our work provides a glean into the distribution of comorbidities and the distribution of health services within the U.S.; coupled with that, however, is also the acknowledgement that there are many factors that may be responsible for this distribution. 

Factors such as state wide funding, federal funding, politics within each state, the mentality towards mental health, are just a few of the factors that contribute to the results we found within our project and data; however, we cannot hope to pinpoint the definitive causality of our results. What we can do is look further into each of these avenues and conduct more research and surveys to deeper contextualize this topic. 

Something we can all do without a smidge of data science knowledge is to aid in reducing the stigma associated with mental health and receiving mental health care. The American Psychological Administration found results from the Harris Poll indicating that young adults, those aged 18-34 (which is a very similar age group to the one we conducted observations on) were less [likely than their older peers](https://www.apa.org/news/press/releases/2019/05/mental-health-survey#:~:text=A%20total%20of%2087%25%20of,better%2C%20according%20to%20the%20poll.) to agree that having a mental health disorder is nothing to be ashamed of.

# References

American Addiction Centers (2022), "Co-Occurring Disorder Treatment Centers Near Me," available at <https://americanaddictioncenters.org/co-occurring-disorders>.

American Addiction Centers (2023), "Mental Health Spending By State Across the US," available at <https://rehabs.com/explore/mental-health-spending-by-state-across-the-us/>.

American Psychological Association (2019), "Survey: Americans Becoming More Open About Mental Health," available at <https://www.apa.org/news/press/releases/2019/05/mental-health-survey#:~:text=A%20total%20of%2087%25%20of,better%2C%20according%20to%20the%20poll>.

Marshall, Lisa (2022), "Multiple diagnoses are the norm with mental illness; new genetic study explains why," available at <https://www.colorado.edu/today/2022/05/10/multiple-diagnoses-are-norm-mental-illness-new-genetic-study-explains-why>.

Mental Health America (2023), "The State Of Mental Health In America", available at <https://mhanational.org/issues/state-mental-health-america>.

Substance Abuse and Mental Health Services Administration (November 2022), "2020 Mental Health Client-Level Data," available at <https://www.samhsa.gov/data/data-we-collect/mh-cld-mental-health-client-level-data>.

Substance Abuse and Mental Health Services Administration (November 2022), "2020 Mental Health Client-Level Data Codebook," available at <https://www.datafiles.samhsa.gov/sites/default/files/field-uploads-protected/studies/MH-CLD-2020/MH-CLD-2020-datasets/MH-CLD-2020-DS0001/MH-CLD-2020-DS0001-info/MH-CLD-2020-DS0001-info-codebook.pdf>.

U.S. Department of Education, "U.S. Department of Education Announces More Than $188 Million...," available at <https://www.ed.gov/news/press-releases/us-department-education-announces-more-188-million-bipartisan-safer-communities-act-support-mental-health-and-student-wellness>. 
 
World Health Organization (2021), "Community-based mental health services using a rights-based approach", available at <https://www.who.int/news-room/feature-stories/detail/community-based-mental-health-services-using-a-rights-based-approach>.

